---
title: "Filtering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

```{r setup}
library(tiledbsc)
library(SeuratObject)
```

# Overview

This vignette demonstrates how to leverage tiledbsc's features for subsetting and filtering data.

## Dimension slicing

### Low-level classes

`TileDBArray` and its subclasses provide a `set_query()` method that allows you to define the ranges of the indexed dimensions to retrieve.

As an example, let's ingest cell-level metadata from the `pbmc_small` dataset into an `AnnotationDataframe` indexed by the cell identifiers.

```{r}
data("pbmc_small", package = "SeuratObject")

adf <- AnnotationDataframe$new(uri = file.path(tempdir(), "pbmc_small_obs"))
adf$from_dataframe(pbmc_small[[]], index_col = "obs_id")
adf
```

As you'd expect, reading the data back into memory recreates the same data frame with all 80 cells.

```{r}
dim(adf$to_dataframe())
```

We can use the `set_query()` method to read in a specific subset of cells by providing a named list of ranges for each dimension.

```{r}
adf$set_query(
  dims = list(obs_id = c("AAATTCGAATCACG", "AAGCAAGAGCTTAG", "AAGCGACTTTGACG"))
)
```

Now, the next time we call `to_dataframe()` the resulting data frame will only contain the cells we specified.

```{r}
adf$to_dataframe()
```

This functionality is also available for `AnnotationGroup`-based classes, which ostensibly contain a group of arrays that all share the same dimensions. In this case, the `dims` argument is automatically applied to all child arrays.

### SOMA

The `SOMA` class also provides a `set_query()` method for the same purpose. However, unlike the lower-level classes which are meant to be general-purpose,
the `SOMA` class has well-defined requirements for the dimensions of its member arrays. As such, `SOMA$set_query()` provides arguments, `obs_ids` and `var_ids`, rather than the more general `dims` argument.

To see this in action, let's create a `SOMA` from `pbmc_small`'s `RNA` assay.

```{r}
soma <- SOMA$new(uri = file.path(tempdir(), "pbmc_small_rna"))
soma$from_seurat_assay(pbmc_small[["RNA"]])
```

Now let's use `set_query()` to slice three features of interest.

```{r}
soma$set_query(var_ids = c("PPBP", "PF4", "GNLY"))
```

Any of the individual SOMA components will respect the new query. So if we read in the feature-level metadata from `var`, the resulting data frame contains only the features we specified.

```{r}
soma$var$to_dataframe()
```

And when we recreate the entire Seurat `Assay` from the `SOMA`, the slice is carried through to all relevant components,

```{r}
pbmc_small_rna <- soma$to_seurat_assay()
```

including the raw counts,

```{r}
GetAssayData(pbmc_small_rna, slot = "counts")
```

and normalized/scaled data:


```{r}
GetAssayData(pbmc_small_rna, slot = "data")
```
