---
title: "Seurat to sc_group"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Seurat to sc_group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

```{r setup}
options(max.print = 500)

library(tiledbsc)
library(fs)
library(tiledb)
library(SeuratObject)

data_dir <- file.path(tempdir(), "pbmc_small")
dir.create(data_dir, showWarnings = FALSE)
```

## Overview

This vignette will cover the creation of an **sc_dataset** from a `Seurat` object.

## Load data

Load the subsetted 10X genomics [PBMC dataset](https://mojaveazure.github.io/seurat-object/reference/pbmc_small.html) provided by SeuratObject.

```{r}
data("pbmc_small", package = "SeuratObject")
pbmc_small
```

## Convert a `Seurat` object to a TileDB-backed `sc_dataset`

The `SCDataset` class provides a method for converting an entire `Seurat` object to an `sc_dataset`. This is the recommended way to perform the conversion since it can handle multiple `Assay` objects and will (eventually) convert all of the standard slots that comprise a `Seurat` object.

This first step is to create a new `SCDataset` object and provide a URI where the dataset should be created:

```{r}
scdataset <- SCDataset$new(uri = file.path(tempdir(), "sc_dataset"))
```

Next, we'll pass the entire `pbmc_small` object directly to `from_seurat()` and one `sc_group` will be created for each `Assay` object:

```{r}
scdataset$from_seurat(object = pbmc_small)
```

Examining the directory structure, you can see the top-level `sc_dataset` directory now contains a single `scgroup_RNA` sub-directory, corresponding to `pbmc_small`'s only assay, `"RNA"`:

```{r}
fs::dir_tree(scdataset$uri, recurse = 1)
```

Internally, the `SCGroup` class is used to convert each Seurat `Assay` object to an `sc_group`, which creates and populates the various sub-components, including:

- the `counts`, `data`, and `scale.data` matrices are each stored in separate attribute of the `X` array
- the `data.frame` containing feature-level metadata is ingested into the `var` array

Separately, any dimensional reductions are extracted and stored in corresponding `obsm`/`varm` arrays.


To close the loop we can then read the on-disk `sc_dataset` back into memory as a `Seurat` object.

```{r}
scdataset$to_seurat(project = "SCDataset Example")
```


## Convert a Seurat `Assay` to TileDB-backed `sc_group`

Conversions can happen at multiple levels of the API. For example, we can operate directly on a Seurat `Assay` using `SCGroup`. The workflow is largely the same:


```{r}
scgroup <- SCGroup$new(uri = file.path(tempdir(), "sc_group"))
fs::dir_tree(scgroup$uri)
```

Then we'll pass `RNA` assay from `pbmc_small` to the `from_seurat_assay()` method of the `SCGroup` class.

*Note: Because cell-level metadata is stored in the parent `Seurat` object, we need to provide this data separately.*

```{r}
scgroup$from_seurat_assay(
  object = pbmc_small[["RNA"]],
  obs = pbmc_small[[]]
)
```

Examine the directory structure of the `sc_group` we can see the `X`, `var`, and `obs` arrays have all been created.

```{r}
fs::dir_tree(scgroup$uri, recurse = FALSE)
```

## Accessing individual components

Any of the underlying TileDB arrays can be accessed directly from an `sc_dataset` object by navigating its internal classes.

As an example, let's access the cell-level metadata. Recall from the [`matrix-api spec`][matrix-api-spec] that cell-level metadata is stored in the `obs` array of an `SCGroup`. Therefore, we must first access a specific `SCGroup` within the `SCDataset`'s `scgroups` slot. Let's generate a list of the available `scgroups`:

```{r}
names(scdataset$scgroups)
```

Easy choice. `"RNA"` can then be used to index the corresponding `SCGroup`:


```{r}
scdataset$scgroups$RNA
```

We can see we have access to a variety of fields and methods, but `obs` is the one we're after.

```{r}
scdataset$scgroups$RNA$obs$to_dataframe()
```

This is a `AnnotationDataframe` object, which includes a method for reading the data into R as a `data.frame`:


```{r}
head(scdataset$scgroups$RNA$obs$to_dataframe())
```


# Array Access

## Helpers

All of the array-based classes include a number of helper functions for interacting with the underlying arrays.

Print the schema of an array:

```{r}
scgroup_obs <- scdataset$scgroups$RNA$obs
scgroup_obs$schema()
```

List the names of the array's dimensions (i.e., indexed columns)

```{r}
scgroup_obs$dimnames()
```

and attributes (i.e., non-indexed columns):

```{r}
scgroup_obs$attrnames()
```

## TileDB API access

You can also use the `tiledb_array()` method to directly access the underlying arrays using the standard TileDB API, providing the full functionality of the [`tiledb`][tiledb-r] package. For example, let's query the `obs` array and  retrieve a subset of cells that match our QC criteria:

```{r}
obs_array <- scgroup_obs$tiledb_array(
  return_as = "tibble",
  attrs = c("nCount_RNA", "nFeature_RNA"),
  query_condition = parse_query_condition(nFeature_RNA < 2500)
)

obs_array[]
```


# Session

```{r}
sessionInfo()
```

```{r cleanup, include=FALSE}
tiledb::tiledb_vfs_remove_dir(data_dir)
```

<!-- links -->
[tiledb-r]: https://tiledb-inc.github.io/TileDB-R/
[matrix-api-spec]: https://github.com/single-cell-data/matrix-api/blob/main/specification.md
