---
title: "Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

This vignette provides a high-level introduction to the tiledbsc R package, intended to get users up and running quickly.

In this tutorial we will cover:

- ingesting a `Seurat` object
- access the individual components
- slicing and filtering
- updating with new results

## Setup

```{r}
library(tiledbsc)
library(SeuratObject)
```

For this tutorial we'll use a dataset from 10X genomics containing 2,700 peripheral blood mononuclear cells (PBMC) that's commonly used in Seurat and Scanpy tutorials.

Use the provided helper function to download the filtered gene / cell matrix from 10X and create a `Seurat` object.

```{r}
pbmc3k <- dataset_seurat_pbmc3k()
pbmc3k
```

## Populate a SOMACollection

Our first step is to create a new `SOMACollection` at a specific URI. The URI could be a local filepath, an S3 URI, or as discussed below, a TileDB Cloud URI which automatically registers the dataset on TileDB Cloud.

```{r}
soco_uri <- file.path(tempdir(), "soco-pbmc3k")

soco <- SOMACollection$new(uri = soco_uri)
soco
```

Now we use tiledbsc to automatically ingest the various components of the `pbmc3k` object into the new `SOMACollection`.

```{r}
soco$from_seurat(pbmc3k)
```

Printing the `soco` object shows it contains 2 members: `RNA`, which contains the RNA assay data, and `uns`, which is a special group for miscellaneous (i.e., unstructured) data.

```{r}
soco
```

## Access SOMA Components

As this dataset is unimodal, we'll extract the SOMA containing the RNA data.

```{r}
soma <- soco$get_member("RNA")
soma
```

Access the [`AnnotationDataframe`] representing the `obs` array containing cell-level metadata stored in `obs`.

```{r}
soma$obs
```

Read it into memory as a `data.frame`.

```{r}
obs <- soma$obs$to_dataframe()
summary(obs)
```

Access the [`AssayMatrix`] representing the `X` array containing the RNA expression data stored in `X`.

```{r}
soma$X$members$counts
```

Read it into memory as a [`dgTMatrix`][`Matrix::TsparseMatrix-class`].

```{r}
mat <- soma$X$members$counts$to_matrix()
summary(mat)
```

## Dimension Slicing and Attribute Filtering

`SOMACollection` and `SOMA` objects can be sliced by a combination of obs and var identifiers.

```{r}
soma$set_query(obs_ids = head(colnames(pbmc3k), 10))
```

This query is automatically applied to the entire SOMA, so any component read into memory will now only return the filtered results.

```{r}
soma$obs$to_dataframe()
```

To remove the filter simply reset the query.

```{r}
soma$reset_query()
nrow(soma$obs$to_dataframe())
```

In addition to dimension slicing we can also apply filters using the cell- and feature-level metadata stored in the `obs` and `var` arrays.

```{r}
soma$set_query(
  obs_attr_filter = nCount_RNA > 3000
)
```

```{r}
pbmc3k[[]]
```


We can load the SOMA slice into memory as a Seurat `Assay`...

```{r}
soma$to_seurat_assay()
```

or as a Bioconductor [`SummarizedExperiment`].

```{r}
if (requireNamespace("SummarizedExperiment", quietly = TRUE)) {
  soma$to_summarized_experiment()
}
```

## Adding Analysis Results to a SOMA

The original `pbmc3k` object contains only the raw counts data. Here we'll use `Seurat` to generate new results and add them to the `soma`.

### Add Layer for Normalized Counts

First we normalize the raw counts, which creates a new matrix stored in the `Assay` object.

```{r eval=requireNamespace("Seurat", quietly = TRUE)}
pbmc3k <- Seurat::NormalizeData(pbmc3k, normalization.method = "LogNormalize")
```

Currently the RNA SOMA's X group contains only a single layer, _counts_.

```{r}
soma$X
```

Add a new `X` layer containing log-normalized counts stored in the Seurat `Assay`'s `data` slot.

```{r}
soma$from_seurat_assay(
    object = pbmc3k[["RNA"]],
    layers = "data",
    var = FALSE
)
```

Now we can see the SOMA contains 2 X layers: `counts` and `data`.

```{r}
soma$X
```

### Add Layer for Scaled counts

First, we identify highly variable subset of genes to scale.

```{r eval=requireNamespace("Seurat", quietly = TRUE)}
var_genes <- VariableFeatures(
  Seurat::FindVariableFeatures(pbmc3k, nfeatures = 2000)
)
```

Then generate the scaled data.

```{r, eval=requireNamespace("Seurat", quietly = TRUE)}
pbmc3k <- Seurat::ScaleData(pbmc3k, features = var_genes)
```


And ingest the scaled data into a new `X` layer called `"scale.data"`.

```{r eval=requireNamespace("Seurat", quietly = TRUE)}
soma$from_seurat_assay(
    object = pbmc3k[["RNA"]],
    layers = "scale.data",
    var = FALSE
)
```

Again, verify the new layer was added.

```{r}
soma$X
```

### Add Dimensional Reduction Results

We can use a similar workflow for dimensional reduction results, which are stored in either the `obsm` or `varm` slots of a SOMA.

```{r eval=requireNamespace("Seurat", quietly = TRUE)}
dimreduc_pca <- Reductions(
  Seurat::RunPCA(pbmc3k, features = var_genes),
  slot = "pca"
)
```

Add the Seurat `DimReduc` object to the SOMA.

```{r eval=requireNamespace("Seurat", quietly = TRUE)}
soma$add_seurat_dimreduction(
    dimreduc_pca,
    technique = "pca"
)
```

The SOMA now contains 2 new arrays, one within the `obsm` group storing the cell embeddings

```{r}
soma$obsm
```

and a second within the `varm` group storing the feature loadings.

```{r}
soma$varm
```

Either of these can be accessed and queried directly, which is often useful for visualization applications.

```{r eval=requireNamespace("Seurat", quietly = TRUE)}
obsm_pca <- soma$obsm$get_member("dimreduction_pca")
obsm_pca
```


```{r eval=requireNamespace("Seurat", quietly = TRUE)}
cell_pcs <- obsm_pca$to_matrix()
pairs(cell_pcs[, 1:5])
```
